### 2020-04-29 manuscript/manuscript.rmd
updates to main

Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life. However, over the past decade, there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig. 1) emerging as the two primary hypotheses. Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. An understanding of the placement of these lineages is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns. However, a wide consensus of animal origin has not been reached despite multiple efforts.

The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative. Since then many more studies have been published, some supporting Ctenophora-sister, some Porifera-sister, and some neither (Table 1). It has become clear that this is a very difficult phylogenetic challenge. As the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem. Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support. Here we synthesize data and results from phylogenomic analyses that tested Ctenophora-sister and Porifera-sister. We reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies. This provides an integrative overview of the challenges posed by this question and provides direction for future studies. We also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.

@@

## Variation across studies

### Models of molecular evolution

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$ , the rate of evolution, and the amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the rate at which one amino acid changes to another (*e.g.* Poisson/F81, WAG or LG, and GTR). The rate of exchange can be further scaled, for example with a Gamma parameter that accommodates rate variation across sites. Heterogeneous equilibrium approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. Models of molecular evolution are assembled by selecting different options for all these different components. The models that are applied in practice are heavily influenced by computational costs, as well as convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). Moreover, the Gamma-rate heterogeneity is used in almost every analysis conducted here. If there are no modifiers, then it is implied that site homogeneous models are used. A detailed description of each component is provided in Supplementary Results.

### Gene and taxa sampling

@@

@@

### Data-recoding

Compositional heterogeneity (clade-specific and site-specific differences in amino acid frequencies) have been a concern in the study of phylogenetics, and has been of particular concern for the deep animal phylogeny [@philippe2011resolving]. Variation in evolutionary rates across taxa can lead to compositional heterogeneity that may lead to long branch attraction of sponges or ctenophores [@King:2017ie]. Feuda *et al*. 2017 [@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in matrices. By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent. However, how the recoding methods impact phylogenetic inference has not been well explored. Moreover, a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six].

>>>>

Historically, there was little debate about the root of the animal tree of life and Porifera-sister (Fig. 1) was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative. Since then many more studies have been published, some supporting Ctenophora-sister, some Porifera-sister, and some neither (Table 1). As the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem. Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support.

@@

## Variation across published studies

The published analyses considere here differ in a variety of ways, including the model of molecular evolution, gene selection, ingroup species selection, and outgroup species selection.

Models of molecular evolution describe how probable different evolutionary changes are, and can have a major impact on the topology of inferred phylogenies. Models have several components that each consider different aspects of the evolutionary process. The models used in the studies considered here differ according to three primary components: the exchangeability matrix $E$ , rates of evolution, and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the rate at which one amino acid changes to another (*e.g.* Poisson/F81, WAG or LG, and GTR). The rate of exchange can be further scaled, for example with a Gamma parameter that accommodates rate variation across sites. Heterogeneous equilibrium approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. Models of molecular evolution are assembled by selecting different options for all these different components. The models that are applied in practice are heavily influenced by computational costs, as well as convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). Moreover, the Gamma-rate heterogeneity is used in almost every analysis conducted here. If there are no modifiers, then it is implied that site homogeneous models are used. A detailed description of each component is provided in Supplementary Results.

@@

Ingroup selection. XXX

@@
